Data-driven Reduction of Multiscale Stochastic Dynamical Systems \thanks{This work was supported by ....}

### 2014-08-30 reducible_SDE/paper/SIADS_draft.tex
added edits after meeting with Ronen

\usepackage{graphicx}
%\usepackage{subcaption}
%\usepackage{epstopdf}

@@

\author{Carmeline J. Dsilva \and Ronen Talmon \and C. William Gear \and Ronald R. Coifman \and Ioannis G. Kevrekidis}

@@

@@

\begin{AMS}\end{AMS}

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{TEX PRODUCTION}{USING SIAM'S MM \LaTeX\ MACROS}

@@

@@

@@

Consider the following system of SDEs:
\begin{equation} \label{eq:fast_slow_SDE}
\begin{aligned}
dx &= adt + dW_1\\
dy &= -\frac{y}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
\end{aligned}
\end{equation}
TODO: add more general formulation of the multi scale SDE

So $x$ is the slow variable, and $y$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
%
We assume that the saturation time of fast scale  of the SDE is smaller than the other relevant timescales.
%
We would like to recover $x$ using data-driven techniques.

@@

\subsection{Nonlinear functions}

We will also consider a (nonlinear) function $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^m$ of the data sampled from the SDE in \eqref{eq:fast_slow_SDE}.

@@

\section{Data mining: Nonlinear Intrinsic Variables}

%These same conditions for uncoupled noises with unit variance will allow us to recover the slow variable(s) from data with multiple timescales. 
%
We want to emphasize the link between multiscale SDEs and NIVs: NIVs rescales the noises, so that the ``fast'' variables become small.
%
Consider the SDE in \eqref{eq:fast_slow_SDE}, and let $y = \frac{z}{\sqrt{\epsilon}}$. 
%
Then
\begin{equation}
\begin{aligned}
dx &= adt + dW_1\\
dz &= -\frac{z}{\epsilon} dt +  dW_2
\end{aligned}
\end{equation}
%
Therefore, $z$ is a stochastic variable whose diffusion is unity, and so $x$ and $z$ are the ``intrinsic variables'' for the underlying SDE.
%
However, the equilibrium measure of $z$ is $\mathcal{O}(\epsilon)$, and so when $\epsilon \ll 1$, $z \ll x$, and so we will recover $x$ (the slow variable). 
%
In general, we want to recover the slow variable(s) even when the observations of the underlying intrinsic stochastic process are obstructed by a {\em nonlinear} measurement function.

We will define $X = \begin{bmatrix} x \\ z \end{bmatrix}$

To get a parameterization of NIV, we require $\| X_2 - X_1\|_2$.
%
However, we only have access to $Y = f(X)$.
%
Assuming that we have a local approximation to $\| X_2 - X_1\|$ from the measurements $Y$, we can then use DMAPS to uncover a parameterization of $X$ from the estimates of the pairwise distances.

@@

@@

\begin{figure}
\includegraphics[width=\textwidth]{schematic.eps}
\caption{Illustration of how to choose $\delta t$ and $\sigma_{kernel}$ appropriately. The data shows the evolution of the ``fast'' variable at a fixed value of the ``slow'' variable.  We must choose both parameters so that the curvature effects and other nonlinearities are negligable. }
\label{fig:schematic}
\end{figure}

@@

@@

To compute the Mahalanobis distance, we require $J J^T$.
%
By Ito formula, we have
\begin{equation}
dY^j = \sum_{i=1}^n \left( \frac{1}{2} f_{ii}^j + a_i f_i^j \right) dt + \sum_{i=1}^n f_i^j dW_i
\end{equation}
%
TODO: define partials...
%
The covariance $C$ defined by 
\begin{equation}
[C]_{jk} = \mathrm{Cov} \left( dY^j, dY^k \right)
\end{equation}
equals $C = J J^T$ \cite{...}.

We will empirically estimate the covariance from data.

Comment: Use a more general model (not necessarily multi scale)

The goal of NIVs is to recover ``intrinsic'' variables that are independent of the measurement/observation function.
%
We define intrinsic variables as those variables which have uncoupled noise with unit variance,
i.e., the intrinsic variables $X^1, \dots, X^n$ are governed by the following SDEs:
\begin{equation} \label{eq:NIV_formulation}
dX^i = a(X) dt + dW^i
\end{equation}
where $X = \begin{bmatrix} X^1 & X^2 & \cdots & X^n \end{bmatrix}^T$.

Restrictions:
\begin{itemize}
\item Drift is bilipshitz wrt process $X$
\item Access to simulator, BUT the analysis could be extended to a single trajectory (cite people....); emphasize the difference between simulation and experiment
\end{itemize}

We assume that
\begin{equation}
 \int_{t}^{t + \delta t} a(X_\tau) d \tau - a(X_t) \delta t  = \mathcal{O} (\delta t) \: \forall t
\end{equation}
TODO: for the toy problems, find the order of this term, possibly as a function of $\epsilon$-- then we can relate $\epsilon$ to $\delta t$, and place bounds on $\delta t$

\begin{equation}
\begin{aligned}
X(t + \delta t) =& X_t + \int_{t}^{t+\delta t} a(X_\tau) d \tau + d W \\
= & X_t + \int_{t}^{t+\delta t} \left[ a(X_t) + \left( a(X_\tau) - a(X_t) \right) \right] d \tau + d W \\
= & X_t + a(X_t) \delta t  + \sqrt{\delta t} W + \mathcal{O} (\delta t)
\end{aligned}
\end{equation}

Given $X$ governed by the SDE in \eqref{eq:NIV_formulation}, we can sample ``bursts'' around a point $X_0$ in order to estimate the local covariance of the noise.

These clouds will be distributed

\begin{equation}
X \sim \mathcal{N}\left( X_0, \delta t I \right)
\end{equation}
where 
$$X_0 = X_t + \left( a(X_t) + \mathcal{O}(1) \right) \delta t $$
%
Note: we bound the random variation due to the variations in the drift, and note that this bound makes these variations smaller than the variations due to the local noise/diffusion. 
%
Therefore, we simply absorb these variations into an $\mathcal{O}(\delta t)$ correction of the mean of the sample distribution, turning a random term into a deterministic term.

Let $f^i(X)$ denote the $i^{th}$ entry of $\mathbf{f}$, and let $f^i_j(X) = \frac{\partial f^i}{\partial X^j}$.
%
By Taylor expansion around $X=X_0$,
%
\begin{eqnarray}
\hat{C}(X_0) &=& \delta t J_1 J_1^T 
+ \frac{\delta t^2}{2} \left[ J_1 J_3^T + J_3 J_1^T  
+ \frac{3}{2} J_2 J_2^T 
-\frac{1 }{2} J_2 1 1^T J_2^T \right]
+ \mathcal{O} (\delta t^3) %\\
%&=& \delta t J_1 J_1^T 
%+ \frac{\delta t^2}{2} U \Sigma V
%+ \mathcal{O} (\delta t^3)
\end{eqnarray}
where
\begin{equation}
\begin{aligned}
J_1(X_0) &= \begin{bmatrix}
f_1^1 & f_2^1 & \dots & f_n^1 \\
\vdots & \vdots & \cdots & \vdots \\
f_1^m & f_2^m & \dots & f_n^m
\end{bmatrix}\\
%
J_2(X_0) &= \begin{bmatrix}
f_{11}^1 & f_{22}^1 & \dots & f_{nn}^1 \\
\vdots & \vdots & \cdots & \vdots \\
f_{11}^m & f_{22}^m & \dots & f_{nn}^m
\end{bmatrix}\\
%
J_3(X_0) &= \begin{bmatrix}
f_{111}^1 & f_{222}^1 & \dots & f_{nnn}^1 \\
\vdots & \vdots & \cdots & \vdots \\
f_{111}^m & f_{222}^m & \dots & f_{nnn}^m
\end{bmatrix}
\end{aligned}
\end{equation}
%
In this equation, we see what we expect (a linear function of the variance, due to the gradient of $f$) and a second (higher-order) term arising from the curvature and third derivative of $f$.
%
The importance of this second term increases as the curvature increases, as well as when $\delta t$ increases.

Then
\begin{equation}
\hat{C}^{-1}(X_0) \approx \frac{1}{\delta t} \left( J J^T \right)^{-1} - A
\end{equation}
%
where 
\begin{equation}
A = \frac{1}{2} (J_1 J_1^T)^{-1} U \left(\Sigma^{-1} + \frac{\delta t}{2} V \left( J_1 J_1^T \right)^{-1} U \right)^{-1} V \left( J_1 J_1^T \right)^{-1}
\end{equation}
and
\begin{equation}
U \Sigma V = J_1 J_3^T + J_3 J_1^T + \frac{3}{2} J_2 J_2^T -\frac{1 }{2} J_2 1 1^T J_2^T 
\end{equation}
$A$ is a function of the first-, second-, and third-order derivatives of $\mathbf{f}$.

@@

From \eqref{eq:taylor_expansion}, we can see that the squared {\em intrinsic} distance $\| X_2 - X_1 \|^2$, is a function of $\|Y_2 - Y_1\|$ and $\delta t$. 
%
If the first term in the expansion dominates, then $\|X_2 - X_1 \|^2$ should be independent of $\delta t$, and should vary quadratically with $\| Y_2 - Y_1\|$. 
%
We can find such a regime empirically (by plotting $\| X_2 - X_1 \|^2$ as a function of $\delta t$ and $\|Y_2 - Y_1\|$), and select a kernel distance where both conditions hold.

@@

\Appendix
\section{The use of appendices}
The \verb|\appendix| command may be used before the final sections
of a paper to designate them as appendices. Once \verb|\appendix|
is called, all subsequent sections will appear as

\appendix
\section{Title of appendix} Each one will be sequentially lettered
instead of numbered. Theorem-like environments, subsections,
and equations will also have the section number changed to a letter.

If there is only {\em one} appendix, however, the \verb|\Appendix|
(with a capital letter) should be used instead. This produces only 
the word {\bf Appendix} in the section title, and does not add a letter.
Equation numbers, theorem numbers and subsections of the appendix
will have the letter ``A''  designating the section number.

If you don't want to title your appendix, and just call it
{\bf Appendix A.} for example, use \verb|\appendix\section*{}| 
and don't include anything in the title field. This works
opposite to the way \verb|\section*| usually works, by including the
section number, but not using a title.

Appendices should appear before the bibliography section, not after,
and any acknowledgments should be placed after the appendices and before
the bibliography.

\begin{thebibliography}{1}
\bibitem{GoMiSa} {\sc M. Goossens, F. Mittelbach, and A. Samarin},
{\em The} \LaTeX\ {\em Companion}, Addison-Wesley, Reading, MA, 1994.

\bibitem{Higham} {\sc N.~J. Higham}, {\em Handbook of Writing for
the Mathematical Sciences}, Society for Industrial and Applied
Mathematics, Philadelphia, PA, 1993.

\bibitem{Lamport} {\sc L. Lamport}, \LaTeX: {\em A Document
Preparation System}, Addison-Wesley, Reading, MA, 1986.

\bibitem{SerLev} {\sc R. Seroul and S. Levy}, {\em A
Beginner's Book of} \TeX, Springer-Verlag, Berlin, New
York, 1991.
\end{thebibliography}

>>>>

\usepackage{epsfig}

@@

\author{Carmeline J. Dsilva\footnotemark[2] \and Ronen Talmon\footnotemark[3] \and C. William Gear\footnotemark[2] \and Ronald R. Coifman\footnotemark[3] \and Ioannis G. Kevrekidis\footnotemark[2]\ \footnotemark[4]}

@@

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\footnotetext[2]{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, 08544, USA}
\footnotetext[3]{Department of Mathematics, Yale University, New Haven, Connecticut, 06520, USA}
\footnotetext[4]{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, 08544, USA}

\renewcommand{\thefootnote}{\arabic{footnote}}

@@

\begin{AMS}
37M10, 62-07
\end{AMS}

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{C.~J. DSILVA {\it ET AL}}{DATA-DRIVEN REDUCTION OF SDES}

@@

\item
Bridge/connect between data mining/analysis and dynamical systems

\item
Main idea and scope: analyze stochastic dynamical systems using data-driven methods.

\item
Reduction in fast-slow systems using manifold learning

\end{itemize}

\end{itemize}
\item
We cannot use off-the-shelf data-driven methods because they are typically based on metrics and geometry and are not informed (take into account) the dynamics and time.

\item
Our contribution:

\item
We show how to use data analysis methods that are typically applied to data sets for analyzing data collected as a time series from a dynamical system. In particular, we ``propose" a data-driven (graph-based) method for recovering the slow variable of reducible systems.

\item
The method is based mostly on local analysis that combines geometry and dynamics.

\item
We offer rigorous analysis for the method that gives the conditions (for our method to successfully recover the ``right" slow variables), intuition.

\end{itemize}

\end{itemize}

@@

Add a general formulation of a reducible system (with a general drift, and decoupled diffusions) and Dror/Kasmiskii's conditions.
(We assume that the saturation time of fast scale  of the SDE is smaller than the other relevant timescales.)
\begin{equation} 
\begin{aligned}
dx &= a(x,y) dt + dW_1\\
dy &= -\frac{b(x,y)}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
\end{aligned}
\end{equation}

@@

We consider systems that can be written as deterministic nonlinear functions $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^m$ of the following system of SDEs (in the general form, in particular, this one):
\begin{equation} \label{eq:fast_slow_SDE}
\begin{aligned}
dx &= adt + dW_1\\
dy &= -\frac{y}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
\end{aligned}
\end{equation}
where the diffusion terms of the fast and slow variables are decoupled.

So $x$ is the slow variable, and $y$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
%
We would like to recover $x$ using data-driven techniques.

@@

@@

\section{Local Invariant Metrics}

Assume we want to locally recover the slow variables. 
%
To do so, we need to build a metric that is invariant to the fast variables. 
%
This is typically done by averaging out the fast variables. 
%
We propose to do this using the Mahalanobis distance, which is a local PCA-type approach that inverts the local covariance matrix of the data, thereby creating a space with normalized diffusion terms.
%
It can be shown that the Mahalanobis distance locally approximates: 
\begin{equation}
\| f(x_1, y_1) - f(x_2, y_2) \|^2_M = \| X_1 - X_2 \|^2_2 + \mathcal{O}(\| f(x_1, y_1) - f(x_2, y_2) \|^4_2)
\end{equation}
where $X$ comprises variables which have uncoupled noise with unit variance,
i.e., the variables $X^1, \dots, X^n$ are governed by the following SDEs:
\begin{equation} \label{eq:NIV_formulation}
dX^i = a(X) dt + dW^i
\end{equation}
where $X = \begin{bmatrix} X^1 & X^2 & \cdots & X^n \end{bmatrix}^T$.
%
In particular for the case TODO, consider the change of variable $z = y\sqrt{\epsilon}$. This change of variables gives the following system:
\begin{equation}
\begin{aligned}
dx &= adt + dW_1\\
dz &= -\frac{z}{\epsilon} dt +  dW_2
\end{aligned}
\end{equation}
Therefore, $z$ is a stochastic variable whose diffusion is unity, and $X = \begin{bmatrix} x \\ z \end{bmatrix}$.

Note that for this specific case
\begin{equation}
\| X_1 - X_2 \|^2_2 = \|x_1 - x_2 \|^2_2 + \epsilon \|y_1 - y_2 \|^2_2.
\end{equation}
Remarkably, it implies two features.
%
One, it rescales that data (in the sense that the variables have unit diffusion terms) so that the fast variables are collapsed and become (``epsilon") small as observed in the second term in the right hand side. 
%
Two, it inverts the function $f$ (locally) and allows for the approximation of the Euclidean distance between samples of the slow variables in the ``decoupled" SDE space.

%These same conditions for uncoupled noises with unit variance will allow us to recover the slow variable(s) from data with multiple timescales. 
%
%

%To get a parameterization of NIV, we require $\| X_2 - X_1\|_2$.
%
%However, we only have access to $Y = f(X)$.
%
%Assuming that we have a local approximation to $\| X_2 - X_1\|$ from the measurements $Y$, we can then use DMAPS to uncover a parameterization of $X$ from the estimates of the pairwise distances.

\section{DMAPS/Graph-Based/Kernel-based method}

@@

\section{Analysis}

\begin{figure}
\epsfig{width=\textwidth, file=schematic.eps}
\caption{Illustration of how to choose $\delta t$ and $\sigma_{kernel}$ appropriately. The data shows the evolution of the ``fast'' variable at a fixed value of the ``slow'' variable.  We must choose both parameters so that the curvature effects and other nonlinearities are negligable. }
\label{fig:schematic}
\end{figure}

See schematic figure.

@@

@@

Given the true covariance, we can write the Euclidean distance $\|X_2 - X_1 \|$ using Taylor expansion as...

Talk about fourth-order error terms, say we want to choose the kernel scale such that these error terms are negligible. 
%
The error terms involve curvature of $f$, etc.

@@

To compute the Mahalanobis distance, we require $J J^T$ (estimate the covariance matrix from the data).
%
We do this using local simulation ``bursts''. 
%
The error in the covariance estimation is... (add equations).
%
The lowest-order error term depends on the curvature of $f$.

@@

TODO: interpret the different terms we get w.r.t. the curvature of f, the drift, the scale epsilon, and the window size.

From \eqref{eq:taylor_expansion}, we can see that the squared distance $\| X_2 - X_1 \|^2$, is a function of $\|Y_2 - Y_1\|$ and $\delta t$. 
%
If the first term in the expansion dominates, then $\|X_2 - X_1 \|^2$ should be independent of $\delta t$, and should vary quadratically with $\| Y_2 - Y_1\|$. 
%
We can find such a regime empirically (by plotting $\| X_2 - X_1 \|^2$ as a function of $\delta t$ and $\|Y_2 - Y_1\|$), and select a kernel distance where both conditions hold.

@@
